Accessing the nucleon tensor structure in inclusive deep inelastic scattering

### 2016-07-27 1607 - JMC for g2/Paper/JMC_dis.tex
Changes, especially in last part

The new term in Eq.\eqref{e:FLTint} only appears in the $g_2$ structure function. Following the derivation in Ref.~\cite{Accardi:2009au}, one finds
\begin{align}
\label{e:g2}
  g_2(x_B) = g_2^{WW} + \frac{1}{2}\,\sum_a e_a^2
\biggl(
    \widetilde g_T^{a \star}(x) 
    + \int_x^1\frac{dy}{y} \widehat{g}_T^q(y) 
    + \frac{\mq}{\Lambda} \left(\frac{h_1^q}{x}\right)^\star(x) 
    + \frac{\mj-\mq}{\Lambda} \frac{h_1^q(x)}{x} 
\Biggr) \ ,
\end{align}
where we defined $f^*(x) = -f(x) + \int_x^1\frac{dy}{y} f(y)$. The first four
terms coincide with the result obtained in the conventional handbag
approximation \cite{Accardi:2009au}, while the fifth is new.

@@

It is important to estimate the size of the various contributions to the non Wandzura-Wilczek part of $g_2$. We define the shorthand notation
\begin{align}
g_2^{\rm tw3} & = \frac{1}{2}\,\sum_a e_a^2
    \biggl(
    \widetilde g_T^{a \star}(x) 
    + \int_x^1\frac{dy}{y} \widehat{g}_T^q(y) 
    \biggr) 
&
g_2^{\text{quark}} &= \frac{1}{2}\,\sum_a e_a^2 
 \frac{\mq}{\Lambda} (h_1^q/x)^\star(x),
&
g_2^{\text{jet}} &= \frac{1}{2}\,\sum_a e_a^2 
\frac{\mj-\mq}{\Lambda} \frac{h_1^q(x)}{x}. 
\end{align} 
These are compared in Figure~\ref{f:g2contrib} to the $g_2-g_2^{WW}$ function
obtained in the very recent JAM15 fit of polarized DIS asymmetries
\cite{Sato:2016tuz}, that includes a large amount of precise data at large $x$
from Jefferson Lab, and simultaneously fits the higher-twist components of in
$g_1$ and $g_2$ to the data. For the ``pure twist-3'' contribution,
$g_2^{\rm tw3}$, {\it i.e.}, the contribution from quark-gluon-quark matrix
elements, we show a model calculation by Braun et al.~\cite{Braun:2011aw}; for
(modified) bag model calculations, see \cite{Jaffe:1990qh,Stratmann:1993aw}. To estimate the contributions
from quark ($g_2^q$) and jet mass ($g_2^{\rm jet}$) effects, that depend on chiral
odd quark-quark matrix elements, we use the recent Pavia15 fit of the
transversity distribution from Ref.~\cite{Radici:2015mwa}, which is comparable
also to other 
extractions~\cite{Anselmino:2013vqa,Kang:2015msa}. Furthermore, we choose the
values of the mass parameters to be $\mq=5$ MeV and $\mj = 100$ MeV.

As one can see, in the proton case the pure twist-3 contribution is quite
smaller in magnitude, and opposite in sign, compared to the JAM15 fit. As
expected, the quark-mass contribution is essentially negligible and cannot
reconcile these two. Even though the uncertainties in the $h_1$ extraction, and
even more in the estimate of $M_q$, are large, it is quite clear that the gap
between the pure twist-3 $g_2^{\rm tw3}$ function and the JAM15 fit can be
explained by the new jet-mass contribution we discussed in this paper.

In the neutron case, the jet contribution is very negative at intermediate to
large values of $x$. If one trusts the order of magnituse of the $g_2^{\rm tw3}$
calculation by Braun et al., one would conclude that the jet contribution
cannot be that large. However, the latter is dominated by the $d$ quark's
transversity, whose fit suffers from large systematic uncertainties and
saturates the negative Soffer bound.Recent deata in $p+p$ collisions indicate,
however, that $h_1^d$ is less negative than in the Pavia15 fits, in agreement
with the JAM15 fit of the non Wandzura-Wilczek contribution to
$g_2$. Correspondingly the jet contribution to the proton at $x \approx 0.1$
would become less positive, inproving as well the agreement with the JAM15
fit.

\begin{itemize}
\item smaller-$x$: constrains the small-$x$ behavior of transversity.
\end{itemize}
Some ideas to check:
\begin{itemize}
\item JAM13 neutron has almost 0 twist 3, with small positive contribution at
  small x. Can we use this to highlight the $g_2^{\rm jet}$ contribution? Or maybe
  as a cnstrain: if Braun is right then $g_2^{\rm jet}(n) \approx
  -g_2^{\rm tw3}(n)$. (still, does not help much for the proton). 
\item In Jam 13, Fig. 6, it looks like in general the violation of the BC sum
  rule is a small-x effect --> constraints on h(x) fits? This cannot really be
  checked vs. JAM15, that assumes $g_2^(0)$ respects BC.  
\end{itemize}

\begin{figure}[tbh]
\begin{center}
\includegraphics[width=8cm]{g2contrib}
\includegraphics[width=8cm]{g2contribN}
\caption{\label{f:g2contrib} 
Different contributions to the non Wandzura-Wilczek part of the proton (left)
and neutron (right) $g_2$ structure function compared to the JAM15 fit of the
$g_2-g_2^{\text{WW}}$ (solid black) \cite{Sato:2016tuz}. The quark and jet
contributions are shown with a dotted red and a dot-dashed green line
respectively, with uncertainty bands coming form the Pavia15 fit of the
transversity function \cite{Radici:2015mwa}. The unceratainty in the choice
$m_q=5$ GeV and $M_q=100$ GeV is not shown. The pure twist-3 contribution
calculated by Braun et al. \cite{Braun:2011aw} is shown as a dashed blue line
(no uncertainty estiamte was provided in the original reference). 
}
\end{center}
\end{figure}

@@

The zero-th moment provides an interesting relationship between the longitudinal and tensor spin structure of the target,
\begin{align}
  \label{eq:BC}
  \int dx\, g_2(x) = \sum_q e_q^2 \frac{\mj-\mq}{\Lambda} \int dx\, \frac{1}{x} h_1^q(x) \ ,
\end{align}
where we used the fact that $\hat g_T^q[0]=0$ identically due to the symmetry properties of the quark-gluon-quark correlators. The sum rule \eqref{eq:BC}, in which the pure twist-3 part does not take part, is a consequence of the Lorentz invariance of QCD interactions, that entails $\int_0^1 dx g_1^a(x) = \int_0^1 g_T^q(x)$. It explicitly displays a contribution from spin-flip processes, that are included in the original derivation by Burkhardt and Cottingham \cite{Burkhardt:1970ti} but do not show up in treatments that only consider free field quark propagators for the struck quark \cite{Jaffe:1996zw}. Since $h_1$ is driven to 0 by QCD evolution as $Q^2 \to \infty$, the BC sum rule $\int_0^1 dx\, g_2(x) =0$ is satisfied at least asymptotically. At finite scales, assuming the Burkhardt-Cottingham sum rule, that is derived at the cross section level (under suitable assumptions on the high-energy behavior of te hspin-flip amplitude) rather than from an analysis of collinear correlators, as done here, to be strictly valid, we obtain a strong constraint on the transversity function,
\begin{align}
   \int dx\, \frac{1}{x} h_1^q(x) = 0 \ .
\label{eq:ABsumrule}
\end{align}
Interestingly, one can show that this constraint, if valid at any given scale $Q_0$ is conserved through QCD evolution. Even if the BC sum rule is broken either by a $J = 0$ fixed pole with non-polynomial residue \cite{Jaffe:1996zw} or explicitly as, \eg, in the quark target model \cite{Kundu:2001pk} -- i.e., if $g_2$ integrates to a finite but nonzero number -- we obtain that $h_1(x)/x$ must be integrable. This entails a bound on the small $x$ behavior of the transversity,
\begin{align}
  h_1^q(x) \propto x^\epsilon \ \ \ \epsilon>0 \ .
\label{eq:ABbound}
\end{align}
This bound will be very useful, \eg, in transversity fits, where the data at
small $x$ is, as yet, very limited, and in general for proper extrapolations
when calculating moments.

>>>>

The new term in Eq.\eqref{e:FLTint} only appears in the $g_2$ structure function. Following the derivation in Ref.~\cite{Accardi:2009au}, one finds
\begin{align}
\label{e:g2}
  g_2(x_B) = g_2^{WW} + \frac{1}{2}\,\sum_a e_a^2
\biggl(
    \widetilde g_T^{a \star}(x) 
    + \int_x^1\frac{dy}{y} \widehat{g}_T^q(y) 
    + \frac{\mq}{\Lambda} \left(\frac{h_1^q}{x}\right)^\star(x) 
    + \frac{\mj-\mq}{\Lambda} \frac{h_1^q(x)}{x} 
\Biggr) \ ,
\end{align}
where we defined $f^*(x) = -f(x) + \int_x^1\frac{dy}{y} f(y)$. The first four
terms coincide with the result obtained in the conventional handbag
approximation \cite{Accardi:2009au}, while the fifth is new. Note that even if
the relation is written for the sum of the quark weighted by their charge
squared, it can be considered valid also flavor by flavor. In fact, the steps
leading to such a decomposition are formulated at the correlator level.

@@

It is important to estimate the size of the various contributions to the non Wandzura-Wilczek part of $g_2$. We define the shorthand notation
\begin{align}
g_2^{\rm tw3} & = \frac{1}{2}\,\sum_q e_q^2
    \biggl(
    \widetilde g_T^{q \star}(x) 
    + \int_x^1\frac{dy}{y} \widehat{g}_T^q(y) 
    \biggr) 
&
g_2^{\text{quark}} &= \frac{1}{2}\,\sum_q e_q^2 
 \frac{\mq}{\Lambda} (h_1^q/x)^\star(x),
&
g_2^{\text{jet}} &= \frac{1}{2}\,\sum_q e_q^2 
\frac{\mj-\mq}{\Lambda} \frac{h_1^q(x)}{x}. 
\end{align} 
These are compared in Figure~\ref{f:g2contrib} to the $g_2-g_2^{WW}$ function
obtained in the very recent JAM15 fit of polarized DIS asymmetries
\cite{Sato:2016tuz}, that includes a large amount of precise data at large $x$
from Jefferson Lab, and simultaneously fits the higher-twist components of in
$g_1$ and $g_2$ to the data. For the ``pure twist-3'' contribution,
$g_2^{\rm tw3}$, {\it i.e.}, the contribution from quark-gluon-quark matrix
elements, we show a model calculation by Braun et al.~\cite{Braun:2011aw}; for
(modified) bag model calculations, see \cite{Jaffe:1990qh,Stratmann:1993aw}. To estimate the contributions
from quark ($g_2^q$) and jet mass ($g_2^{\rm jet}$) effects, that depend on chiral
odd quark-quark matrix elements, we use the recent Pavia15 fit of the
transversity distribution from Ref.~\cite{Radici:2015mwa}, which is comparable
also to other 
extractions~\cite{Anselmino:2013vqa,Kang:2015msa}. Furthermore, we choose the
values of the mass parameters to be $\mq=5$ MeV and $\mj = 100$ MeV.

As one can see, in the proton case the pure twist-3 contribution is quite
smaller in magnitude, and opposite in sign, compared to the JAM15 fit. As
expected, the quark-mass contribution is essentially negligible. 
For what concerns the jet-mass contribution, the uncertainties due to the $h_1$
extraction are very large, especially at low $x$. In addition, there is an
overall normalization uncertainty due to the choice of $\mj$, not shown in the
plot. In any case, the jet-mass contribution is strikingly large. If we assume
that the pure twist-3 contributions are of the order of the model calculation
by Braun et al., the breaking of the Wandura-Wilczek relation can be used to
constrain the extractions of the transversity distribution, in particular at
low $x$. 
Moreover, it is quite clear that the gap
between the pure twist-3 $g_2^{\rm tw3}$ function and the JAM15 fit can be
explained by the new jet-mass contribution we discussed in this paper.

In the neutron case, the jet contribution is very negative at intermediate to
large values of $x$. 
If one trusts the order of magnitude of the $g_2^{\rm tw3}$
calculation by Braun et al., one would conclude that the jet contribution
should not be that large. However, the jet contribution is strongly influenced
by the $d$ quark's
transversity, whose fit suffers from large systematic uncertainties and
saturates the negative Soffer bound. Recent deata in $p+p$ collisions indicate,
however, that $h_1^d$ might be less negative than in the Pavia15
fits~\cite{Radici:2016lam}.
Correspondingly the jet contribution to the proton at $x \approx 0.1$
would become less positive, inproving as well the agreement with the JAM15
fit.

\begin{figure}[tbh]
\begin{center}
\includegraphics[width=8cm]{g2contrib}
\includegraphics[width=8cm]{g2contribN}
\caption{\label{f:g2contrib} 
Different contributions to the non Wandzura-Wilczek part of the proton (left)
and neutron (right) $g_2$ structure function compared to the JAM15 fit of the
$g_2-g_2^{\text{WW}}$ (solid black) \cite{Sato:2016tuz}. The quark and jet
contributions are shown with a dotted red and a dot-dashed green line
respectively, with uncertainty bands coming form the Pavia15 fit of the
transversity function \cite{Radici:2015mwa}. The unceratainty in the choice
$m_q=5$ GeV and $M_q=100$ GeV is not shown. The pure twist-3 contribution
calculated by Braun et al. \cite{Braun:2011aw} is shown as a dashed blue line
(no uncertainty estimate was provided in the original reference). 
}
\end{center}
\end{figure}

@@

The zero-th moment provides an interesting relationship between transversity
and the inclusive structure function $g_2$:
\begin{align}
  \label{eq:BC}
  \int dx\, g_2(x) = \sum_q e_q^2 \frac{\mj-\mq}{\Lambda} \int dx\, \frac{1}{x} h_1^q(x) \ ,
\end{align}
where we used the fact that $\hat g_T^q[0]=0$ identically due to the symmetry
properties of the quark-gluon-quark correlators. 
The sum rule \eqref{eq:BC}
generalizes the Burkhardt-Cottingham (BC) sum rule~\cite{Burkhardt:1970ti},
which states that  $\int_0^1 dx\, g_2(x) =0$. 
However, we show that jet-mass corrections violate the BC sum rule. The
possibility of a violation of the sum rule due to 
contributions from spin-flip
processes was already mentioned in the original
derivation~\cite{Burkhardt:1970ti}, 
but do not show up in treatments that only
consider free field quark propagators for the struck quark
\cite{Jaffe:1996zw}. Since $h_1$ is driven to 0 by QCD evolution as $Q^2 \to
\infty$, the BC sum rule $\int_0^1 dx\, g_2(x) =0$ is satisfied at least
asymptotically. Although we formulated \eqref{eq:BC} in terms of sum over quark flavors in order to display a clear
connection to the structure function $g_2$, we stress that it is valid also
flavor by flavor, i.e., for each single flavor the only measurable nonzero contribution to the zeroth moment of the
structure function $g_2$ can come from the jet-mass
corrections and transversity.\footnote{This conclusion is true even if the BC sum rule is broken by a $J = 0$ fixed pole with
non-polynomial residue \cite{Jaffe:1996zw}, since this would appear as a
$\delta(x)$ contribution and would not be measurable.}
The fact that $g_2^{\text{quark}}$ disappears in the zeroth moment
of $g_2$ is due to the definition of the $\ast$ functions. The fact that $g_2^{\rm tw3}$
disappears is due to the symmetry properties of $\widehat{g}_T^q$, or
equivalently  is a consequence of the
Lorentz invariance of QCD interactions, that entails $\int_0^1 dx g_1^a(x) =
\int_0^1 g_T^q(x)$.

At finite scales, the only way to preserve the validity of the
Burkhardt-Cottingham sum rule is if
%that is derived at the cross section level (under suitable assumptions on the
%high-energy behavior of te hspin-flip amplitude) rather than from an analysis
%of collinear correlators, as done here, 
%to be strictly valid, we obtain a
%strong constraint on the transversity function, 
\begin{align}
   \int dx\, \frac{1}{x} h_1^q(x) = 0 \ .
\label{eq:ABsumrule}
\end{align}
Interestingly, one can show that this constraint, if valid at any given scale
$Q_0$ is conserved through QCD evolution. However, we think that this
constrain cannot be satisfied in general, since it is broken in perturbative
QCD~\cite{Kundu:2001pk} and models (see, e.g., \cite{}).

E or explicitly as, \eg, in the quark
target model \cite{Kundu:2001pk} -- i.e., if $g_2$ integrates to a finite but
nonzero number -- 
If we assume that the BC sum rule is broken by a {\em finite} amount, we
obtain that $h_1(x)/x$ must be integrable, implying a 
a bound on the small $x$ behavior of the transversity, 
\begin{align}
  h_1^q(x) \propto x^\epsilon \ \ \ \epsilon>0 \ .
\label{eq:ABbound}
\end{align}
This bound will be very useful, \eg, in transversity fits, where the data at
small $x$ is, as yet, very limited, and in general for proper extrapolations
when calculating moments.
